WED-SQL (Draft)

### 2016-05-20 doc/paper/main_en.tex
paper: ch4

Most of modern business process management software systems are subject to recurrent structural modifications during its
life cycle. This changes are necessary to incorporate new requisites to the software, once it is often impossible 
to come up with all requirements upfront in the modeling phase of project - some of then only appear after the software
is running in production. Furthermore, it is too expensive and time consuming to try to predict all the possibles situations 
and its outcomes, which ultimately may raise exceptions that must be dealt with. Over time, these modifications leads
to code deterioration, exponentially increasing the maintenance costs and compromising the quality of the system as well 
negatively impacting its overall performance.
\\
\indent  Classic business process specification models, for example WS-BPEL, Process Algebra and Petri Networks, focus primarily
on behavioral interactions and relations between process, pushing data analysis into the background and overlooking its
relevance for the project. Therefore, none of these models exactly fits the needs of design dynamic business process 
management systems, which may result in costly system adaptations and maintenance.
\\
\indent  The WED-flow approach, in contrast with the classical models, capture both the inter-process relations and, with the very
same relevance, the data generated by their interactions. With this in mind, the WED-flow allows systems architects to 
think in terms of data states and transitions, similarly to a finite automata designing. Modifying or adding new business 
rules to the project, which is often translated in creation of new processes, is a simple matter of defining new data states 
and conditions for transitions, thus capturing the true nature of dynamic processes and simplifying an incremental evolution 
of the whole project. Real time exceptions are also easier to be handled and integrated to the project, once this approach
is based on transactional properties, thus providing several recovery mechanisms~\cite{ICWS12}.
\\
%problemas da implementacao e justificativa do WED-SQL
%tolerancia a falhas
\indent In order to provide a standard implementation as well as simplifying its software development, in this work we present 
the WED-SQL, an WED-flow framework. The main purpose of WED-SQL is to implement all WED-flow definitions~\cite{FTPM10} and encapsulate 
the control structure, what includes the transactional control, exceptions handling, data states detection, state transitioning 
and ensure data integrity. This framework effectively is an abstraction layer that allow developers to put more effort
on business logic and come up with better WED-flow projects, also making better use of reusable code. 
\\
\indent The WED-SQL framework has been built inside the PostgreSQL~\cite{PSQL} relational database management system (PostgreSQL RDBMS)
, thereby uses SQL language
to specify WED-flow elements, has full transactional support and it is highly fault tolerant. Once it employs well consolidate 
technologies, widespread in the computer science field, this framework targets to disseminate the adoption of WED-flow on
business process management software design, by means of an easy to use, reliable and scalable tool that is capable to 
provide the flexibility demanded by modern transaction control systems. 
\\
\indent The remaining of this paper is structured as follows: Section II presents WED-SQL general software architecture aspects and
internal data structures. Section III briefly describes how the system works. A discussion about the algorithms involved
is presented in section IV. Following, a detailed description of transactional control and the communication protocol used
for external communication is given in section V. Finally, on section VI, we discuss the next challenges to improve and
expand this framework capabilities.

@@

To better understand what the WED-SQL does, first one needs to understand the basic definitions of the WED-flow paradigm. 
A WED-flow application is composed of a set of WED-attributes $\mathcal{A} = \{a_1,a_2,a_3,\ldots,a_n\}$, a set of WED-states
$\mathcal{S} = \{s_1,s_2,s_3,\ldots,s_m\}$ where $ \forall s_i \in \mathcal{S}, s_i = (v_1,v_2,v_3,\ldots,v_n)$
is a n-tuple where each $i\in[1,n], v_i$ is a value for $a_i \in \mathcal{A}$ and the size of $s_i = |\mathcal{A}|$, a set of
WED-conditions $\mathcal{C}$, a set of WED-transitions $\mathcal{T}$ and a set of WED-triggers $\mathcal{G}$. A WED-condition
$c \in \mathcal{C}$ is a predicate over $\mathcal{S}$, that is, we say that $s \in \mathcal{S}$ satisfies $c$ if the values
of $s$ makes $c$ true. A WED-transition is a function $t: \mathcal{S}\rightarrow\mathcal{S} \in \mathcal{T}$ that receives 
as input a WED-state $s$ and returns another WED-state $s'$. Finally, a WED-trigger $g = (c,t)$ where $g \in \mathcal{G},
c \in \mathcal{C}, t \in \mathcal{T}$ is a 2-tuple that associates a WED-condition $c$ with a WED-transition $t$. Therefore,
an instance of a WED-flow application stars with a initial WED-state $s$ that should match some WED-condition $c$ to fire
the WED-trigger $g$ that, in turn, will initiate a WED-transition $t$, which must complete by some predefined timeout,thus 
generating the next WED-state. This keeps going until a final WED-state is reached, when this instance is then finalized. 
Roughly speaking, a WED-flow application is a transactional data state transitioning machine.
\\
\indent
Aiming to be capable of running in a distributed environment, for instance in a web-services composition way, as well
to broadly support Long Lived Transaction handling, the WED-SQL is based on a Client-Server system architecture and has
two main component thereafter: a server component called WED-server and a client component called WED-worker. The
sever side is responsible for both transactional and flow control, what means that besides coordinating the WED-transitions, 
enforcing individually defined completion time limits, it is also in charge of matching WED-states against WED-conditions 
and WED-trigger firing. On the other hand, the client side is responsible for doing the computation needed to perform each
WED-transition. 
\\
\subsection{WED-worker and WED-server}
Fundamentally, the WED-server is a PostgreSQL extension package consisting of \emph{database triggers}, control tables and 
stored procedures~\cite{NAV}. Choosing PostgreSQL as the WED-server foundation was a decision based on several of its
aspects, such as be an open source project, catalog-driven operation, support for user-written code using low and high level 
programming languages (e.g., C, Python, Perl, TCL and SQL itself).
\\
\indent  WED-SQL directly benefits from the fact that, being an Open-source software, PostgreSQL's license states that it can be used, 
modified, copied and distributed without any kind of payment or agreement. Besides that, having access to the source code
makes development less complex and error prone once it is easier to analyze and, consequently, better comprehend its internal
mechanisms. 
\\
\indent Relational database systems needs to store metadata about internal structure of tables, columns, indexes, etc. This data
is kept on what is know as system catalogs. PostgreSQL not only stores much of its control data in these catalogs
but also base its procedures on then, in what is called a catalog-driven operation. Since they are readily  
available to users as ordinary tables, thus can be modified, it is fairly easy to modify the PostgreSQL behavior. This
property is specially useful to allow the WED-server to manage timed transactions related to WED-transitions and exception 
handling. 
\\
\indent As previously mentioned, users can write PostgreSQL modules using a few different languages. The most simple and straight
forward of then is a SQL dialect called pgSQL, which is basically SQL with variables and flow control structures. Also, modules
can be written in C language and loaded as a dynamic library, providing the user with ultimate logic control and performance.
Despite these two language having native support in PostgreSQL, sometimes pgSQL is not expressive enough when the logic
involved is too complex, while C, albeit fast, is too low level and overloaded with technical implementation details too
perform most of the tasks. The very PostgreSQL documentation recommends using a language other than C unless it is explicit
needed. Fortunately, it also offers support for Python, a much more versatile and expressive programming language than
the two former. Given the dynamic nature of WED-flow models, Python has been proven the perfect tool to do most of the job
on the WED-server side, only recurring to C when critical or where an extra performance is absolutely necessary. And,of
course, SQL statements are every where in both WED-server and WED-workers. 
\\
\indent The role of the WED-workers is to perform the WED-transitions. They must connect to the WED-server and "ask" for a pending
job before opening a new transaction. All WED-transitions are encapsulated in a single transaction with a time limit to 
complete, otherwise they are automatically aborted by the WED-server. Each WED-transition must have at least one WED-worker
associated and, depending on the workload on the server side, can have many more. On the other hand, each WED-worker is 
specialized in execute a specific WED-transition. The maximum number of WED-workers simultaneously transacting depends 
on the number of concurrent connections that the WED-server is able to keep alive.

@@

\end{figure}
 The diagram depicted on figure* represents the data model used by the WED-server. It is not a true Entity-Relationship 
model once the relation between the \emph{WED\_attr} and \emph{WED\_flow} (represented by the arrow in the diagram) is 
managed by a stored procedure rather than by the RDBMS. This is necessary to comply with the dynamic nature of the WED-flow 
approach, which allows the system designer to evolve the system on-the-fly, for instance, including new WED-attributes and 
WED-conditions. In that case, the system's data structure may need to be modified accordingly. 
\\
\indent WED-attributes are represented by the entity \emph{WED\_attr} using the following attributes:
\begin{itemize}
\item $aname$: Name of the WED-attribute and also the primary key;
\item $adv$: Default value for this WED-attribute.
\end{itemize}

The \emph{WED\_flow} entity represents the instances of a WED-flow applications. Besides the identifying attribute (\emph{wid}),
it has a extra one for each WED-attribute defined for a given application, in other words, for each element of \emph{WED\_attr}
there is an attribute $aname$ in \emph{WED\_flow}.
\\
\indent The weak entities \emph{Job\_Pool} and \emph{WED\_trace} results from the relation between the entities \emph{WED\_flow} and 
\emph{WED\_trig}. Identified by the ternary relationship \emph{Match}, \emph{Job\_Pool} represents pending WED-transitions, that is,
the ones awaiting to be processed by a WED-worker, using the following attributes:

@@

A new WED-flow application is created by defining a set of WED-attributes and a set of WED-triggers that associates a 
WED-condition to a WED-transition. This is accomplished, for the time being, writing these definition in SQL (see an example
on Fig.~\ref{code_new}. In the foreseen future we plan to propose a SQL language extension to express at least the basic WED-flow operations. 
\\
\indent Note that the expression of a WED-condition predicate (\emph{cpred}) uses the same syntax of the SQL clause \emph{WHERE}.
In fact, this expression is employed \emph{as-is} for the WED-server while matching it against a WED-state. When surrounded
by double \$ marks, special symbols like quotes don't need to be escaped in the predicate.
\\
\indent  Also note that, although the application's final condition statement is declared in the same WED\_trig table it does not
have a WED-transition associated, otherwise it would be ignored by the WED-server anyway. To declare a WED-condition as
final, one only need to specify its predicate and set the value of the \emph{cfinal} attribute to true. If the final condition
is missing, all instances of a given application will end up in an exception state. 
\\
\indent We encourage developers to encapsulate all definitions for a new WED-flow application into a single transaction. By doing
so one can prevent a partially defined application in case of syntax errors, which may force the developer to manually truncate
the system tables before running the script again. 
\\
\indent Although it is possible to define multiple applications in a single WED-flow database, the WED-server allows each one of
then to be isolated in a different database. Therefore, applications can be grouped together by some semantic meaning and WED-attributes
that must not be shared between some of then can be isolated.

@@

try:
    w.run()
except KeyboardInterrupt:
    print()
    sys.exit(0)
\end{Verbatim}
\caption{Standard WED-worker implementation in Python}
\label{fig_ww}
\end{figure}
The first step to write a WED-worker using the BaseWorker package is to import the module \emph{BaseClass}, which in fact
is an abstract class that implements the communication protocol with the WED-server as well as manages all connections
and transactions involved.  
\\
\indent  Next, a concrete class must implement the abstract method \emph{wed\_trans()} from BaseClass and initialize the following
class attributes: 
\begin{itemize}
\item \emph{trname}: WED-transition name that will be carried out by this WED-worker. Must match the name used on the WED-flow definition;
\item \emph{dbs}: WED-server connection parameters conforming the \emph{psycopg2} driver format. At least the username, database name and
                 the WED-worker name must be provided;
\item \emph{wakeup\_interval}: Time interval in which the WED-worker execution is suspended (sleeping) if there is no new WED-transitions
                              to be performed (explained in details on section~\ref{not}); 
\end{itemize}

The wed\_trans() method works on a specific WED-flow instance at a time. It receives, as a parameter, the WED-state that 
fired this WED-transition and returns a string representing a new WED-state. This return value must comply with the same
syntax used to specify a \emph{SET} clause of a \emph{UPDATE} SQL statement. An empty value can be returned to abort the
transaction.
\\
\indent Finally, the concrete class can be instantiated and executed by invoking its \emph{run()} method.

@@

When a new instance is initiated or when a WED-worker successfully commits a new WED-state, the WED-server henceforth runs 
its main algorithm (Alg.~\ref{alg1}). It starts by evaluating all WED-conditions predicates against the new WED-state. Then, for
each satisfied predicate, the WED-transition associated with this WED-condition will be fired and registered on the Job\_Pool table
,a notification will also be sent to the respective WED-worker. Besides that, this events will be added to the instance's
tracing record on the WED\_trace table. If a WED-state satisfies the final condition and if there is no pending WED-transitions,
either running or awaiting in the Job\_Pool table, for a given instance, it is then marked as final and cannot be further
modified. In the case when there is no pending WED-transitions and the new WED-state is not final, the instance will be
marked as an exception and a special record will be inserted into Job\_Pool to signal some external recovery mechanism~\cite{ICWS12}.
It is worth mentioning that unless a initial WED-state is also final, it must fire 
at least one WED-transition, otherwise the WED-server will reject the new instance. 
\\
\indent  WED-workers spend most of their idle time suspended, waiting for a notification send by the WED-server when there is work 
to be done. After some time waiting, if not notification has arrived they wake up and ask the WED-server for work (this
behavior will be explained in section~\ref{not}). Either way, before initiate a WED-transition, the WED-worker need to 
"inform" the WED-server of in which instance and what WED-transition it is about to perform. This is done by requesting the
WED-server a special lock. If this lock is granted, then the WED-worker can proceed (See alg.~\ref{alg2})
\\
\indent This lock system is necessary once more than one WED-worker can be simultaneously activated for the same WED-transition.
Therefore, its main role is to avoid the same WED-transition from being performed for the same WED-flow instance at the
same time. Furthermore, it also helps the WED-server to manage transactions. It is wise to assume that no WED-transition
can be committed by a WED-worker not holding a lock.
\\
\indent Once in possession of a lock, the WED-worker must finish its job within the time limit defined for the WED-transition.
The transaction is successfully finalized updating the current WED-state for this instance on the WED\_flow table.

@@

\section{Transactional Management}
\label{GT}
Cada instância de um WED-flow pode ser vista como uma Transação Longa (LLT) e, de acordo com ~\cite{SGD87},
suas WED-transitions podem ser vistas como passos Saga. Sendo assim, é função do WED-server garantir a integridade dos estados
de dados, manter a consistência transacional e garantir que todo WED-flow termine ou em um estado final ou em um estado
de exceção e, nesse caso, deve permitir que um passo compensatório possa ser executado. 
\par
Nas tópicos seguintes serão apresentados em detalhes os mecanismos de controle transacional empregados e o protocolo de
comunicação entre múltiplos WED-workers e um WED-server.

%DETALHES DE FUNCIONAMENTO (seriação,locks(),keep connection alive(pessimistic locking),excessoes, tras simultaneas,continuous query,protocolo de comunicaocao, balanço: coneccoes x demanda,payload pode nao ser o estado atual mvcc)
\subsection{Notificações}
\label{not}
%send notify vs continuous query
Ao modelar um sistema cujas operações principais são realizadas com base em eventos de dados, é preciso também conceber-se
algum tipo de mecanismo capaz de capturar tais eventos. Quando esses dados estão armazenados em um SGBD, uma possível 
solução é utilizar um modelo conhecido por \emph{Continuous Query}~\cite{CQ}. Esse método consiste em definir 
uma consulta, com parâmetros específicos para o estado de dados a ser detectado, e executá-la a cada intervalo de tempo.
\par Ao utilizar esse modelo em um sistema sensível ao tempo de resposta na detecção de eventos, é preciso considerar seu
custo computacional que depende tanto da quantidade de eventos distintos quanto da frequência em que ocorrem. Muitos
eventos distintos implica em muitas consultas distintas. Para detectar eventos dos quais não é possível ter uma estimativa razoável
da frequência em que ocorrem, é necessário executar essas consultas no menor intervalo de tempo possível afim de não comprometer  
a capacidade do sistema de responder a eventos em tempo real. Isso tudo acarreta em um considerável desperdício de recursos
computacionais, aumentando o consumo de energia e sobrecarregando o sistema.
\par Dado que o WED-SQL é também um sistema de captura de estados de dados em tempo real, onde o WED-server já realiza
essa tarefa, será que não é possível aproveitar essa computação e notificar cada WED-worker da ocorrência de seu evento
de interesse ? Sim, é possível ! 
\par O PostgreSQL oferece um mecanismo que pode ser utilizado como uma alternativa ao modelo \emph{Continuous Query}, as Notificações Assíncronas.
Por meio do comando \emph{NOTIFY <canal>,<payload>}, o SGDB pode notificar de modo assíncrono um cliente que tenha se registrado, via comando 
\emph{LISTEN <canal>}, em um canal de notificação específico. Além disso, também é possível enviar uma mensagem ao cliente, por meio de "<payload>". Essa mensagem pode
ser, por exemplo, uma cadeia de caracteres no formato \emph{JSON} que represente o estado de dados de uma instância de um WED-flow.
\par A grande vantagem do modelo de Notificações Assíncronas sobre o modelo de Continuous Query é que o cliente recebe a notificação em tempo real
da ocorrência do evento de dados, contanto que o mesmo esteja escutando no canal de notificação nesse momento. Além disso, devido ao fato de o cliente
não precisar verificar a cada certo intervalo de tempo se há alguma nova mensagem, ele pode esperar pela notificação "dormindo", ou seja, sua execução
pode ser suspensa enquanto não há uma nova mensagem, liberando os recursos da máquina para realizar outras tarefas assim como economizando energia.
\par No WED-SQL são utilizados os dois modelos de detecção de eventos de dados. Durante uma execução normal, o WED-server notifica os WED-workers
sempre que uma nova WED-trigger é disparada, ou seja, sempre que surge uma nova tarefa na tabela Job\_Pool. Além disso, a mensagem enviada aos WED-workers
é exatamente a nova tarefa que foi gerada, eliminando a necessidade de se fazer uma busca na tabela Job\_Pool. Cada WED-transition possui um canal exclusivo
de notificações identificado por seu nome definido na tabela WED\_trig. Os WED-workers registram-se em seu respectivo canal e suspendem sua operação até
que sejam acordados com uma notificação. Caso uma notificação seja enviada e não haja ninguém escutando em um determinado canal, a mesma será descartada.
\par Afim de garantir que todas as WED-transitions sejam eventualmente executadas, e de preferência sejam iniciadas no menor intervalo de tempo possível, os

>>>>

Most of modern business process management software systems are subject to recurrent structural modifications during its
life cycle. This changes are necessary to incorporate new requisites to the software, once it is often impossible 
to come up with all requirements upfront in the modeling phase of project - some of then only appear after the software
is running in production. Furthermore, it is too expensive and time consuming to try to predict all the possibles situations 
and its outcomes, which ultimately may raise exceptions that must be dealt with. Over time, these modifications leads
to code deterioration, exponentially increasing the maintenance costs and compromising the quality of the system as well 
negatively impacting its overall performance.

\par  Classic business process specification models, for example WS-BPEL, Process Algebra and Petri Networks, focus primarily
on behavioral interactions and relations between process, pushing data analysis into the background and overlooking its
relevance for the project. Therefore, none of these models exactly fits the needs of design dynamic business process 
management systems, which may result in costly system adaptations and maintenance.

\par  The WED-flow approach, in contrast with the classical models, capture both the inter-process relations and, with the very
same relevance, the data generated by their interactions. With this in mind, the WED-flow allows systems architects to 
think in terms of data states and transitions, similarly to a finite automata designing. Modifying or adding new business 
rules to the project, which is often translated in creation of new processes, is a simple matter of defining new data states 
and conditions for transitions, thus capturing the true nature of dynamic processes and simplifying an incremental evolution 
of the whole project. Real time exceptions are also easier to be handled and integrated to the project, once this approach
is based on transactional properties, thus providing several recovery mechanisms~\cite{ICWS12}.

%problemas da implementacao e justificativa do WED-SQL
%tolerancia a falhas
\par In order to provide a standard implementation as well as simplifying its software development, in this work we present 
the WED-SQL, an WED-flow framework. The main purpose of WED-SQL is to implement all WED-flow definitions~\cite{FTPM10} and encapsulate 
the control structure, what includes the transactional control, exceptions handling, data states detection, state transitioning 
and ensure data integrity. This framework effectively is an abstraction layer that allow developers to put more effort
on business logic and come up with better WED-flow projects, also making better use of reusable code.

\par The WED-SQL framework has been built inside the PostgreSQL~\cite{PSQL} relational database management system (PostgreSQL RDBMS)
, thereby uses SQL language
to specify WED-flow elements, has full transactional support and it is highly fault tolerant. Once it employs well consolidate 
technologies, widespread in the computer science field, this framework targets to disseminate the adoption of WED-flow on
business process management software design, by means of an easy to use, reliable and scalable tool that is capable to 
provide the flexibility demanded by modern transaction control systems.

\par The remaining of this paper is structured as follows: Section II presents WED-SQL general software architecture aspects and
internal data structures. Section III briefly describes how the system works. A discussion about the algorithms involved
is presented in section IV. Following, a detailed description of transactional control and the communication protocol used
for external communication is given in section V. Finally, on section VI, we discuss the next challenges to improve and
expand this framework capabilities.

@@

To better understand what the WED-SQL does, first one needs to understand the basic definitions of the WED-flow paradigm. 
A WED-flow application is composed of a set of WED-attributes $\mathcal{A} = \{a_1,a_2,a_3,\ldots,a_n\}$, a set of WED-states
$\mathcal{S} = \{s_1,s_2,s_3,\ldots,s_m\}$ where $ \forall s_i \in \mathcal{S}, s_i = (v_1,v_2,v_3,\ldots,v_n)$
is a n-tuple where each $i\in[1,n], v_i$ is a value for $a_i \in \mathcal{A}$ and the size of $s_i = |\mathcal{A}|$, a set of
WED-conditions $\mathcal{C}$, a set of WED-transitions $\mathcal{T}$ and a set of WED-triggers $\mathcal{G}$. A WED-condition
$c \in \mathcal{C}$ is a predicate over $\mathcal{S}$, that is, we say that $s \in \mathcal{S}$ satisfies $c$ if the values
of $s$ makes $c$ true. A WED-transition is a function $t: \mathcal{S}\rightarrow\mathcal{S} \in \mathcal{T}$ that receives 
as input a WED-state $s$ and returns another WED-state $s'$. Finally, a WED-trigger $g = (c,t)$ where $g \in \mathcal{G},
c \in \mathcal{C}, t \in \mathcal{T}$ is a 2-tuple that associates a WED-condition $c$ with a WED-transition $t$. Therefore,
an instance of a WED-flow application stars with a initial WED-state $s$ that should match some WED-condition $c$ to fire
the WED-trigger $g$ that, in turn, will initiate a WED-transition $t$, which must complete by some predefined timeout,thus 
generating the next WED-state. This keeps going until a final WED-state is reached, when this instance is then finalized. 
Roughly speaking, a WED-flow application is a transactional data state transitioning machine.

\par
Aiming to be capable of running in a distributed environment, for instance in a web-services composition way, as well
to broadly support Long Lived Transaction handling, the WED-SQL is based on a Client-Server system architecture and has
two main component thereafter: a server component called WED-server and a client component called WED-worker. The
sever side is responsible for both transactional and flow control, what means that besides coordinating the WED-transitions, 
enforcing individually defined completion time limits, it is also in charge of matching WED-states against WED-conditions 
and WED-trigger firing. On the other hand, the client side is responsible for doing the computation needed to perform each
WED-transition.

\subsection{WED-worker and WED-server}
Fundamentally, the WED-server is a PostgreSQL extension package consisting of \emph{database triggers}, control tables and 
stored procedures~\cite{NAV}. Choosing PostgreSQL as the WED-server foundation was a decision based on several of its
aspects, such as be an open source project, catalog-driven operation, support for user-written code using low and high level 
programming languages (e.g., C, Python, Perl, TCL and SQL itself).

\par  WED-SQL directly benefits from the fact that, being an Open-source software, PostgreSQL's license states that it can be used, 
modified, copied and distributed without any kind of payment or agreement. Besides that, having access to the source code
makes development less complex and error prone once it is easier to analyze and, consequently, better comprehend its internal
mechanisms.

\par Relational database systems needs to store metadata about internal structure of tables, columns, indexes, etc. This data
is kept on what is know as system catalogs. PostgreSQL not only stores much of its control data in these catalogs
but also base its procedures on then, in what is called a catalog-driven operation. Since they are readily  
available to users as ordinary tables, thus can be modified, it is fairly easy to modify the PostgreSQL behavior. This
property is specially useful to allow the WED-server to manage timed transactions related to WED-transitions and exception 
handling.

\par As previously mentioned, users can write PostgreSQL modules using a few different languages. The most simple and straight
forward of then is a SQL dialect called pgSQL, which is basically SQL with variables and flow control structures. Also, modules
can be written in C language and loaded as a dynamic library, providing the user with ultimate logic control and performance.
Despite these two language having native support in PostgreSQL, sometimes pgSQL is not expressive enough when the logic
involved is too complex, while C, albeit fast, is too low level and overloaded with technical implementation details too
perform most of the tasks. The very PostgreSQL documentation recommends using a language other than C unless it is explicit
needed. Fortunately, it also offers support for Python, a much more versatile and expressive programming language than
the two former. Given the dynamic nature of WED-flow models, Python has been proven the perfect tool to do most of the job
on the WED-server side, only recurring to C when critical or where an extra performance is absolutely necessary. And,of
course, SQL statements are every where in both WED-server and WED-workers.

\par The role of the WED-workers is to perform the WED-transitions. They must connect to the WED-server and "ask" for a pending
job before opening a new transaction. All WED-transitions are encapsulated in a single transaction with a time limit to 
complete, otherwise they are automatically aborted by the WED-server. Each WED-transition must have at least one WED-worker
associated and, depending on the workload on the server side, can have many more. On the other hand, each WED-worker is 
specialized in execute a specific WED-transition. The maximum number of WED-workers simultaneously transacting depends 
on the number of concurrent connections that the WED-server is able to keep alive.

@@

\end{figure}
 The diagram depicted on figure* represents the data model used by the WED-server. It is not a true Entity-Relationship 
model once the relation between the \emph{WED\_attr} and \emph{WED\_flow} (represented by the arrow in the diagram) is 
managed by a stored procedure rather than by the RDBMS. This is necessary to comply with the dynamic nature of the WED-flow 
approach, which allows the system designer to evolve the system on-the-fly, for instance, including new WED-attributes and 
WED-conditions. In that case, the system's data structure may need to be modified accordingly.

\par WED-attributes are represented by the entity \emph{WED\_attr} using the following attributes:
\begin{itemize}
\item $aname$: Name of the WED-attribute and also the primary key;
\item $adv$: Default value for this WED-attribute.
\end{itemize}

The \emph{WED\_flow} entity represents the instances of a WED-flow applications. Besides the identifying attribute (\emph{wid}),
it has a extra one for each WED-attribute defined for a given application, in other words, for each element of \emph{WED\_attr}
there is an attribute $aname$ in \emph{WED\_flow}.

\par The weak entities \emph{Job\_Pool} and \emph{WED\_trace} results from the relation between the entities \emph{WED\_flow} and 
\emph{WED\_trig}. Identified by the ternary relationship \emph{Match}, \emph{Job\_Pool} represents pending WED-transitions, that is,
the ones awaiting to be processed by a WED-worker, using the following attributes:

@@

A new WED-flow application is created by defining a set of WED-attributes and a set of WED-triggers that associates a 
WED-condition to a WED-transition. This is accomplished, for the time being, writing these definition in SQL (see an example
on Fig.~\ref{code_new}. In the foreseen future we plan to propose a SQL language extension to express at least the basic WED-flow operations.

\par Note that the expression of a WED-condition predicate (\emph{cpred}) uses the same syntax of the SQL clause \emph{WHERE}.
In fact, this expression is employed \emph{as-is} for the WED-server while matching it against a WED-state. When surrounded
by double \$ marks, special symbols like quotes don't need to be escaped in the predicate.

\par  Also note that, although the application's final condition statement is declared in the same WED\_trig table it does not
have a WED-transition associated, otherwise it would be ignored by the WED-server anyway. To declare a WED-condition as
final, one only need to specify its predicate and set the value of the \emph{cfinal} attribute to true. If the final condition
is missing, all instances of a given application will end up in an exception state.

\par We encourage developers to encapsulate all definitions for a new WED-flow application into a single transaction. By doing
so one can prevent a partially defined application in case of syntax errors, which may force the developer to manually truncate
the system tables before running the script again.

\par Although it is possible to define multiple applications in a single WED-flow database, the WED-server allows each one of
then to be isolated in a different database. Therefore, applications can be grouped together by some semantic meaning and WED-attributes
that must not be shared between some of then can be isolated.

@@

try:
    w.run()
except KeyboardInterrupt:
    print()
    sys.exit(0)
\end{Verbatim}
\caption{Standard WED-worker implementation in Python}
\label{fig_ww}
\end{figure}
The first step to write a WED-worker using the BaseWorker package is to import the module \emph{BaseClass}, which in fact
is an abstract class that implements the communication protocol with the WED-server as well as manages all connections
and transactions involved.

\par  Next, a concrete class must implement the abstract method \emph{wed\_trans()} from BaseClass and initialize the following
class attributes: 
\begin{itemize}
\item \emph{trname}: WED-transition name that will be carried out by this WED-worker. Must match the name used on the WED-flow definition;
\item \emph{dbs}: WED-server connection parameters conforming the \emph{psycopg2} driver format. At least the username, database name and
                 the WED-worker name must be provided;
\item \emph{wakeup\_interval}: Time interval in which the WED-worker execution is suspended (sleeping) if there is no new WED-transitions
                              to be performed (explained in details on section~\ref{not}); 
\end{itemize}

The wed\_trans() method works on a specific WED-flow instance at a time. It receives, as a parameter, the WED-state that 
fired this WED-transition and returns a string representing a new WED-state. This return value must comply with the same
syntax used to specify a \emph{SET} clause of a \emph{UPDATE} SQL statement. An empty value can be returned to abort the
transaction.

\par Finally, the concrete class can be instantiated and executed by invoking its \emph{run()} method.

@@

When a new instance is initiated or when a WED-worker successfully commits a new WED-state, the WED-server henceforth runs 
its main algorithm (Alg.~\ref{alg1}). It starts by evaluating all WED-conditions predicates against the new WED-state. Then, for
each satisfied predicate, the WED-transition associated with this WED-condition will be fired and registered on the Job\_Pool table
,a notification will also be sent to the respective WED-worker. Besides that, this events will be added to the instance's
tracing record on the WED\_trace table. If a WED-state satisfies the final condition and if there is no pending WED-transitions,
either running or awaiting in the Job\_Pool table, for a given instance, it is then marked as final and cannot be further
modified. In the case when there is no pending WED-transitions and the new WED-state is not final, the instance will be
marked as an exception and a special record will be inserted into Job\_Pool to signal some external recovery mechanism~\cite{ICWS12}.
It is worth mentioning that unless a initial WED-state is also final, it must fire 
at least one WED-transition, otherwise the WED-server will reject the new instance.

\par  WED-workers spend most of their idle time suspended, waiting for a notification send by the WED-server when there is work 
to be done. After some time waiting, if not notification has arrived they wake up and ask the WED-server for work (this
behavior will be explained in section~\ref{not}). Either way, before initiate a WED-transition, the WED-worker need to 
"inform" the WED-server of in which instance and what WED-transition it is about to perform. This is done by requesting the
WED-server a special lock. If this lock is granted, then the WED-worker can proceed (See alg.~\ref{alg2})

\par This lock system is necessary once more than one WED-worker can be simultaneously activated for the same WED-transition.
Therefore, its main role is to avoid the same WED-transition from being performed for the same WED-flow instance at the
same time. Furthermore, it also helps the WED-server to manage transactions. It is wise to assume that no WED-transition
can be committed by a WED-worker not holding a lock.

\par Once in possession of a lock, the WED-worker must finish its job within the time limit defined for the WED-transition.
The transaction is successfully finalized updating the current WED-state for this instance on the WED\_flow table.

@@

\section{Transactional Management}
\label{GT}

Each instance of a WED-flow application can be treated as a Long Lived Transaction (LLT), thereby, according to ~\cite{SGD87},
the WED-transitions may be regarded as \emph{SAGA} steps. Therefore, it is a WED-server obligation to ensure the integrity
of data states by means of enforcing the transactional consistence, and thus guarantee that every instance always end in
a final WED-state. In the case where an exception is raised, it must provide support for an external recovery mechanism,
eventually leading this instance to a final WED-state.

\par In the following topics we discuss in details how the WED-server manages the timed transactions. We also present how
to tackle some concurrency control challenges as well as explain the WED-SQL communication protocol.

%DETALHES DE FUNCIONAMENTO (seriação,locks(),keep connection alive(pessimistic locking),excessoes, tras simultaneas,continuous query,protocolo de comunicaocao, balanço: coneccoes x demanda,payload pode nao ser o estado atual mvcc)
\subsection{Notifications}
\label{not}
%send notify vs continuous query

Event-driven computer systems can be roughly described as a system that reacts in the face of a new event. In order to 
accomplish that, event detection mechanisms are often integrated to such systems. Since a data event based system usually keeps
its data stored in a database, data event detection is generally made by a polling operation, that is, by continually checking
some data for changes. This database polling technique, known as \emph{Continuous Query}~\cite{CQ}, mainly consists of 
continuously run a hand tailored query for each targeted data event.

\par When employing the Continuous Query model in a system that is response time sensitive, one needs to keep in mind the
computational cost of this approach, which in turn depends both on the number of distinct events to be detected and on how 
often they pop up. In order to implement this model efficiently, it is critical to have a good time estimation of events
occurrence, especially in the presence of multiple distinct events, once there is an overhead cost involved in performing
a query that cannot be disregarded. In the other hand, if this time estimation is unpredictable for some or all of the
events, the suitable query must be executed as frequent as possible, otherwise threatening the system's real time response 
capability. Anyway, the Continuous Query model is not well suitable to every data event driven system, which ultimatelly
may lead to waste of computational resources, increasing the overall power consumption and possible overloading the system.

\par Since the WED-SQL is also a real time data event detection system, it is indeed part of the WED-server main algorithm,
isn't possible to take advantage of this and somehow inform the WED-workers of the occurrence of its event of interest ?
The answer is yes ! It can.

\par The PostgreSQL provides a mechanism that can be used as an alternative to the Continuous Query model, it is named 
\emph{Asynchronous Notification}. Via the command \emph{NOTIFY}, the DBMS can asynchronously notify a client listening that 
is listening to a communication channel. A clients, in turn, can registry itself via command \emph{LISTEN} on a specific  
channel to receive notifications. Furthermore, the server can also send the client a payload message attached to each
notification.

\par The biggest advantage of the Asynchronous Notification model over Continuous Query is that, providing that a client is 
listening to a given communication channel, messages can be delivered instantly after the detection of a event. Moreover,
now the client doesn't need to poll the server, so it can be suspended until a new notification arrives, thus freeing 
computational resources as well as saving power. In the WED-SQL framework, these two mechanisms are employed in a complementary
combination.

\par Under normal execution, WED-workers are notified whenever a new WED-trigger is fired, in other words, this happens 
every time that a new record is inserted into the Job\_Pool table. In addition, the payload message sent together with
the notification is exactly the new "job" generated, so the WED-workers do not need to go to the WED-server to fetch a task,
saving time and the cost of an extra query. Every WED-transition, in turn, has a proper notification channel identified
by its own name as recorded on the WED\_trig table. WED-workers subscribe themselves to his WED-transition channel and 
go into suspended mode until they are awakened with a new notification. In case a notification is sent and there is nobody
listening to that channel, it will be discarded.

\par Once the WED-server and the WED-workers could be running on physically different machines, communicating over a network, 
messages can get lost. In order to ensure that all WED-transitions are eventually performed as well as to minimize their waiting to run time,
that is, the time that a WED-transition sits into the Job\_Pool table waiting to be performed, WED-workers don't rely solely
on asynchronous notifications. In fact, before they even register on a notification channel, their first step is to scan
the Job\_Pool table for pending WED-transitions, and only then eventually goes into suspension state. With this in mind,
WED-workers can be also configured to wake up after some time suspended and go check for available tasks on the WED-server.
